---
title: "fastPLS Public API and Implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastPLS Public API and Implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`fastPLS` exposes a compact public API. Users do not need to call the internal
C++, CUDA, or benchmarking helpers directly; all model families and backends are
selected through the user-facing functions below.

| Function | Purpose |
|---|---|
| `pls()` | Fit PLSSVD, SIMPLS, OPLS, or kernel PLS for regression or classification. |
| `predict()` | Predict from fitted `fastPLS`, OPLS, or kernel PLS models. |
| `plot()` | Plot PLS or PCA score maps, with optional confidence or Hotelling T2 ellipses. |
| `pls.single.cv()` | Run fixed-component grouped cross-validation. |
| `optim.pls.cv()` | Choose the best number of components by cross-validation. |
| `pls.double.cv()` | Run nested/double cross-validation with inner component optimization. |
| `fastsvd()` | Run a stand-alone truncated SVD through the bundled backends. |
| `pca()` | Compute PCA using the same SVD backend selector. |
| `fastcor()` | Compute fast Pearson-style correlations. |
| `ViP()` | Compute variable-importance-in-projection trajectories. |
| `has_cuda()` | Check whether CUDA-native fastPLS support is available. |
| `has_metal()` | Check whether Apple Metal-native fastPLS support is available. |

## Example Data

The examples below use small simulated data so the vignette is fast to build.
Regression responses are supplied as numeric vectors or matrices. Classification
responses are supplied as factors; internally, `fastPLS` handles the response
encoding and returns predicted class labels.

```{r}
library(fastPLS)

set.seed(100)
n <- 80
p <- 12
X <- matrix(rnorm(n * p), nrow = n, ncol = p)

y_signal <- 0.8 * X[, 1] - 0.5 * X[, 2] + rnorm(n, sd = 0.25)
Y_reg <- matrix(y_signal, ncol = 1)
Y_cls <- factor(ifelse(y_signal > median(y_signal), "high", "low"))

test_id <- sample(seq_len(n), 16)
Xtrain <- X[-test_id, , drop = FALSE]
Xtest <- X[test_id, , drop = FALSE]
Ytrain_reg <- Y_reg[-test_id, , drop = FALSE]
Ytest_reg <- Y_reg[test_id, , drop = FALSE]
Ytrain_cls <- Y_cls[-test_id]
Ytest_cls <- Y_cls[test_id]
```

## Fit PLS Models

The `method` argument selects the outer PLS algorithm. The `backend` argument
selects the implementation: `"cpu"` for the compiled CPU implementation,
`"cuda"` for GPU-native execution when available, and experimental `"metal"`
for Apple Metal acceleration on Apple Silicon.

The `svd.method` argument selects the CPU singular-vector backend for CPU
models. Backend-specific controls are supplied through `...` using the same
compact names as `fastsvd()`, for example `oversample`, `power`, `maxit`,
`tol`, `svtol`, and `seed`.

```{r}
fit_reg <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:3,
  method = "simpls",
  backend = "cpu",
  svd.method = "irlba",
  fit = TRUE
)

fit_reg$Q2Y
```

The same call can fit the four model families by changing `method`:

- `method = "plssvd"`: direct cross-covariance SVD PLS.
- `method = "simpls"`: accelerated compiled SIMPLS implementation.
- `method = "opls"`: OPLS filtering followed by the selected inner PLS core.
- `method = "kernelpls"`: kernel PLS with linear, radial-basis, or polynomial kernels.

## Implementation of the Main Methods

### Singular-vector backends

The PLS algorithms in `fastPLS` are separated from the inner singular-vector
backend. The compiled CPU implementation supports two truncated SVD routes:
`irlba` and randomized SVD. The `irlba` backend follows the implicitly
restarted Lanczos bidiagonalization strategy of Baglama and Reichel (2005). It
estimates only the leading singular triplets needed by the requested PLS
components, rather than computing a full dense SVD. The randomized SVD backend
follows the randomized range-finder framework described by Halko, Martinsson,
and Tropp (2011): it draws a Gaussian test matrix, forms a low-dimensional
sketch of the target matrix, optionally applies power iterations,
orthonormalizes the sketch, and solves the final SVD in the small projected
space. In `pls()` and the cross-validation helpers, these controls are supplied
directly through `...`; in stand-alone SVD calls they are supplied directly to
`fastsvd()`.

On Apple Silicon builds where `has_metal()` returns `TRUE`,
`fastsvd(backend = "metal", method = "rsvd")` and
`pca(backend = "metal", method = "rsvd")` can use the Metal backend. This backend keeps the same
randomized SVD mathematics as the CPU backend, but sends the large sketching matrix
multiplications to Apple Metal Performance Shaders. The QR decomposition and
final small SVD remain on the CPU. Because matrices are copied between R and
Metal buffers, this backend is most useful for larger matrices where the GPU
multiplication work is large enough to dominate transfer overhead.

The experimental PLS Metal backend, `pls(..., backend = "metal")`, extends the
same idea to PLSSVD, SIMPLS, OPLS, and kernel PLS. The large cross-products,
score projections, deflation matrix products, and prediction matrix products
are dispatched through Metal. Small decompositions, scalar reductions, and some
kernel-centering operations remain CPU-side until dedicated Metal kernels are
added.

For CUDA builds, GPU execution is selected with `backend = "cuda"`, so the main
matrix operations are performed by GPU-native PLS routines. When the
cross-covariance \(S = X^T Y\) would be large, selected backends can use the
matrix-free `xprod` route. In that route, the code avoids explicitly storing
`S` and evaluates the products required by the SVD through identities such as
\(Sv = X^T(Yv)\) and \(S^Tu = Y^T(Xu)\). For classification problems with many
classes, label-aware products can avoid materializing a dense one-hot response
matrix. The low-rank streamed prediction path is inspired by the FlashSVD
streaming low-rank inference design of Shao et al. (2025), but is adapted here
to the PLS prediction factors exposed by `fastPLS`.

### PLSSVD

`method = "plssvd"` implements the direct cross-covariance SVD formulation of
PLS, within the same broad PLS regression and PLS-DA framework described by
Wold, Sjostrom, and Eriksson (2001), Barker and Rayens (2003), and Boulesteix
and Strimmer (2007). After preprocessing, the algorithm computes the dominant
singular subspace of \(X^T Y\) and reuses that subspace to generate the
requested component path.
The maximum meaningful number of PLSSVD components is capped by
`min(nrow(X), ncol(X), ncol(Y))`; if a larger value is requested, the package
uses the valid internal rank and records the capped component count.

The main innovation relative to a simple textbook PLSSVD implementation is the
execution strategy. fastPLS can use truncated IRLBA or randomized SVD, compact
low-rank prediction factors, CUDA-native fitting, and the matrix-free `xprod`
route. For large classification tasks, the label-aware PLSSVD path computes
class-wise cross-products from the factor labels and avoids dense one-hot `Y`
unless the legacy dense route is explicitly needed.

### SIMPLS

`method = "simpls"` is the optimized fastPLS SIMPLS core. Statistically, it
keeps the de Jong (1993) SIMPLS idea: one supervised latent direction is
appended at a time, the score and loading vectors are computed, an orthogonal
loading direction is constructed, and the cross-covariance is deflated before
the next component.

The implementation differs from a naive classical SIMPLS loop in how it obtains
new singular directions. Instead of always calling the SVD backend for one fresh
vector after every deflation, fastPLS can refresh a small block of candidate
directions from the current deflated cross-covariance. The block is consumed
sequentially: each candidate direction is normalized, converted into a SIMPLS
component, orthogonalized, and followed by the usual SIMPLS deflation before the
next component is accepted. The cached-deflation update explicitly computes
\(v^T S\) and applies the rank-one update \(S <- S - v(v^T S)\), reducing
temporary allocations and repeated projection work. The incremental refresh
uses the previous latent direction as a warm start for the next randomized
refresh, which can reduce the number of iterations needed to recover the next
dominant direction.

These block-refresh, warm-start, and cached-deflation optimizations are
SIMPLS-specific because SIMPLS is a sequential deflation method. PLSSVD uses a
different one-shot cross-covariance SVD strategy. OPLS and kernel PLS reuse the
same optimized compiled building blocks where their internal fitting step is
linear PLS-like.

### OPLS

`method = "opls"` follows the orthogonal PLS idea introduced by Trygg and Wold
(2002). It first removes response-orthogonal variation from `X` and then fits
the optimized PLS core to the filtered matrix. The number of orthogonal
components is controlled by `north`. The filtering step estimates orthogonal
scores and loadings, subtracts the corresponding variation from `X`, and stores
the orthogonal filter so the same transformation is applied at prediction time.
Users select OPLS with `method = "opls"`; the lower-level implementation details
are handled internally.

### Kernel PLS

`method = "kernelpls"` follows the kernel PLS formulation of Rosipal and Trejo
(2001). It constructs a kernel representation before applying the selected
inner PLS model. The supported kernels are `"linear"`, `"rbf"`, and `"poly"`.
For nonlinear kernels, the training kernel matrix is centered and the PLS model
is fitted in kernel space. For the linear kernel, fastPLS avoids unnecessary
kernel materialization and uses the ordinary linear PLS path.

### Classification heads

For classification, factor responses are treated as PLS-DA targets. The default
classifier, `classifier = "argmax"`, predicts the class with the largest
predicted dummy-response score. This is the classical PLS-DA decision rule
described in chemometric discrimination work such as Barker and Rayens (2003)
and reviewed by Boulesteix and Strimmer (2007).

The LDA head, `classifier = "lda"`, fits a
regularized linear discriminant model on the PLS latent scores, following the
linear discriminant principle of Fisher (1936). For each class, fastPLS
estimates a score-space mean, an empirical prior, and a common pooled
within-class covariance matrix. A small ridge term scaled by the average
covariance diagonal is added for numerical stability. Prediction uses the
standard LDA discriminant score and returns the largest-scoring class. The CUDA
and Metal routes are selected automatically from `backend` when available.

The cKNN head, `classifier = "cknn"`, is the short public name for the
candidate-kNN classifier used in large multiclass PLS-DA.
It first scores classes by centroids in supervised PLS score space, keeps the
top candidate classes, and then reranks those candidates by same-class nearest
neighbours in the PLS score space, using the nearest-neighbour classification
principle of Cover and Hart (1967) after a supervised PLS compression step. The
default tuning parameters are
`candidate_knn_k = 10`, `candidate_tau = 0.2`, `candidate_alpha = 0.75`, and
`candidate_top_m = 20`. `predict()` can also return ranked predictions by using
`top` or `top5 = TRUE`.

```{r}
fit_cls <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpu",
  svd.method = "rsvd",
  seed = 100
)

head(fit_cls$Ypred)
```

For PLS-DA with an LDA prediction head, use the `classifier` argument. The
recommended GPU route is PLSSVD followed by native CUDA LDA:

```{r eval = FALSE}
fit_cls_lda_gpu <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 50,
  method = "plssvd",
  backend = "cuda",
  classifier = "lda"
)
```

When CUDA is unavailable, the compiled CPU fallback is:

```{r}
fit_cls_lda_cpu <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpu",
  svd.method = "rsvd",
  seed = 100,
  classifier = "lda"
)

head(fit_cls_lda_cpu$Ypred)
```

## OPLS and Kernel PLS

OPLS and kernel PLS are accessed through the same `pls()` function. Use
`method = "opls"` for orthogonal filtering before PLS, and
`method = "kernelpls"` for the kernel representation.

```{r}
fit_opls <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:2,
  method = "opls",
  north = 1,
  svd.method = "rsvd",
  seed = 101
)

class(fit_opls)
```

Kernel PLS changes the representation of the samples before the inner PLS fit.
The `linear` kernel is equivalent to an ordinary inner-product representation
and is useful as a fast baseline. The `rbf` kernel uses a radial-basis
similarity and can capture smooth nonlinear class boundaries; its scale is
controlled by `gamma`. The `poly` kernel uses polynomial interactions among
features and is controlled by `degree`, `gamma`, and `coef0`.

```{r}
kernel_fits <- lapply(c("linear", "rbf", "poly"), function(k) {
  pls(
    Xtrain, Ytrain_cls, Xtest, Ytest_cls,
    ncomp = 1:2,
    method = "kernelpls",
    kernel = k,
    gamma = 1 / ncol(Xtrain),
    degree = 2,
    coef0 = 1,
    svd.method = "rsvd",
    seed = 102
  )
})
names(kernel_fits) <- c("linear", "rbf", "poly")

kernel_accuracy <- vapply(kernel_fits, function(fit) {
  mean(fit$Ypred[, ncol(fit$Ypred)] == Ytest_cls)
}, numeric(1))
kernel_accuracy
```

## Prediction

`predict()` applies the preprocessing stored in the fitted object. If `Ytest` is
supplied, predictive metrics are returned when available. The optional
`backend` argument can select the ordinary CPU path, the compact
FlashSVD-style CPU path, CUDA prediction on CUDA-enabled builds, or Metal
prediction on Apple Metal-enabled builds.

```{r}
pred_reg <- predict(fit_reg, Xtest, Ytest = Ytest_reg, proj = TRUE)
names(pred_reg)
dim(pred_reg$Ypred)
```

```{r}
pred_cls <- predict(fit_cls, Xtest, Ytest = Ytest_cls)
table(pred_cls$Ypred[[2]], Ytest_cls)
```

## Score Plots

`plot()` is an S3 method for both fitted PLS models and PCA objects. For PLS
models, refit with `fit = TRUE` to store training scores, or predict with
`proj = TRUE` to store test scores. Ellipses can be ordinary confidence
ellipses or Hotelling T2 ellipses, following the multivariate confidence
region idea introduced by Hotelling (1931).

```{r fig.width = 5, fig.height = 4}
plot(fit_reg, groups = Ytrain_cls, ellipse = TRUE, ellipse.type = "confidence")
```

The next example uses the bundled two-class colon gene-expression data. The
same `plot()` call works for `method = "plssvd"`, `"simpls"`, `"opls"`, and
`"kernelpls"`; here we keep the code short and show one fitted model.

```{r colon-binary-pls-plots, fig.width = 5.5, fig.height = 4.5}
data("colon", package = "fastPLS")
set.seed(200)
X_two <- as.matrix(colon$X)
y_two <- colon$y
train_two <- unlist(tapply(seq_along(y_two), y_two, function(i) {
  sample(i, floor(0.7 * length(i)))
}))

fit_two <- pls(
  X_two[train_two, , drop = FALSE], y_two[train_two],
  X_two[-train_two, , drop = FALSE], y_two[-train_two],
  ncomp = 1:2, method = "simpls", backend = "cpu",
  svd.method = "rsvd", fit = TRUE, proj = TRUE, seed = 200
)

plot(fit_two, score.set = "train", groups = y_two[train_two],
     ellipse = TRUE, main = "Colon, two classes")
```

The second example uses a bundled three-subtype breast-cancer expression
example and fits OPLS with two orthogonal components (`north = 2`). The example
uses predefined training and test sets.

```{r breast-tcga-three-class-opls, fig.width = 7.5, fig.height = 4}
data("breast", package = "fastPLS")
X_three <- as.matrix(breast$X_train)
y_three <- breast$y_train
X_three_test <- as.matrix(breast$X_test)
y_three_test <- breast$y_test

opls_three <- pls(
  X_three,
  y_three,
  X_three_test,
  y_three_test,
  ncomp = 1:2,
  method = "opls",
  backend = "cpu",
  svd.method = "rsvd",
  fit = TRUE,
  proj = TRUE,
  north = 2L,
  seed = 201
)

old_par <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 0.8))
plot(
  opls_three,
  score.set = "train",
  groups = y_three,
  ellipse = TRUE,
  ellipse.type = "hotelling",
  main = "OPLS training, north = 2"
)
plot(
  opls_three,
  score.set = "test",
  groups = opls_three$Ypred[[length(opls_three$ncomp)]],
  xlim = c(-0.16, 0.42),
  main = "OPLS test prediction"
)
par(old_par)
```

## Cross-Validation

`fastPLS` provides three cross-validation helpers. Use `pls.single.cv()` when
the number of latent components is fixed before validation. Use
`optim.pls.cv()` when the number of components should be selected from a grid.
Use `pls.double.cv()` for nested validation, where an inner CV chooses the
number of components and an outer CV estimates predictive performance. All
three helpers support regression and classification and dispatch to the same
compiled CPU/CUDA core when available.

### Ordinary k-fold CV

In ordinary k-fold CV, samples are split directly into `kfold` folds. The
example below validates a fixed two-component SIMPLS classifier with five folds.

```{r}
cv_kfold <- pls.single.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  ncomp = 2,
  kfold = 5,
  method = "simpls",
  backend = "cpu",
  svd.method = "rsvd",
  oversample = 10L,
  power = 1L,
  seed = 103
)

cv_kfold$metrics
```

### Grouped k-fold CV with `constrain`

The `constrain` argument controls grouped splitting. It is a vector with one
entry per sample; samples with the same value are assigned to the same fold. In
practice, this prevents leakage when multiple rows come from the same patient,
subject, batch, or technical replicate. For example, if two spectra come from
the same patient, giving them the same patient identifier in `constrain` ensures
that both spectra are placed either in the training set or in the test set,
never one in each.

```{r}
patient_id <- rep(seq_len(nrow(Xtrain) / 2), each = 2)

cv_grouped <- pls.single.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  constrain = patient_id,
  ncomp = 2,
  kfold = 4,
  method = "simpls",
  backend = "cpu",
  svd.method = "rsvd",
  seed = 104
)

c(n_folds = length(unique(cv_grouped$fold)),
  n_patient_groups = length(unique(patient_id)))
```

### Leave-one-out CV and leave-one-group-out CV

Leave-one-out CV is requested with `kfold = "loocv"`. When `constrain` is not
supplied, each sample is held out once. When `constrain` is supplied, LOOCV
becomes leave-one-constraint-group-out CV, so a whole patient, subject, batch,
or replicate group is held out together at each iteration. Numeric `kfold`
values greater than or equal to the number of constraint groups are also treated
as leave-one-group-out CV.

```{r}
cv_loocv <- pls.single.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  constrain = patient_id,
  ncomp = 1,
  kfold = "loocv",
  method = "simpls",
  backend = "cpu",
  svd.method = "rsvd",
  seed = 105
)

c(n_loocv_folds = length(unique(cv_loocv$fold)),
  n_patient_groups = length(unique(patient_id)))
```

### Component optimization CV

`optim.pls.cv()` repeats the same CV splitting strategy over several candidate
component counts and returns the best value according to the predictive metric:
accuracy for classification, Q2 for univariate regression, and RMSD-style
prediction error for multivariate regression outputs.

```{r}
cv_opt <- optim.pls.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_reg,
  ncomp = 1:3,
  kfold = 5,
  method = "plssvd",
  svd.method = "irlba",
  maxit = 1000L
)

cv_opt$optim_comp
```

### Nested or double CV

`pls.double.cv()` separates model selection from performance estimation. The
inner loop selects the best number of components, while the outer loop evaluates
the selected model on held-out data. Both `kfold_inner` and `kfold_outer` can be
ordinary fold counts or `"loocv"`, and both respect `constrain`.

```{r}
cv_double <- pls.double.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  constrain = patient_id,
  ncomp = 1:2,
  runn = 1,
  kfold_inner = 3,
  kfold_outer = 3,
  method = "simpls",
  backend = "cpu",
  svd.method = "rsvd",
  oversample = 10L,
  power = 1L,
  seed = 104
)

cv_double$bcomp
```

## PCA and SVD Utilities

`fastsvd()` provides direct access to the same SVD backends used by PLS.
`pca()` builds PCA scores and loadings from those backends, following the
classical principal-component construction of Pearson (1901).

```{r}
s <- fastsvd(Xtrain, ncomp = 3, backend = "cpu", method = "rsvd", seed = 104)
length(s$d)
```

```{r fig.width = 5, fig.height = 4}
pc <- pca(Xtrain, ncomp = 3, backend = "cpu", method = "rsvd", seed = 105)
plot(pc, groups = Ytrain_cls, ellipse = TRUE)
```

## Gaussian Response Compression

For high-dimensional responses, `gaussian_y = TRUE` compresses `Y` to a smaller
random Gaussian response space before fitting and decodes predictions back to
the original response scale or class labels. This option uses the random
projection idea underlying the Johnson-Lindenstrauss lemma (Johnson and
Lindenstrauss, 1984) as a practical response-side compression device. By
default, the number of Gaussian dimensions is `min(ncol(Xtrain), 100)` unless
`gaussian_y_dim` is supplied.

```{r}
Y_multi <- cbind(Ytrain_reg, Ytrain_reg^2, sin(Ytrain_reg))
Y_multi_test <- cbind(Ytest_reg, Ytest_reg^2, sin(Ytest_reg))

fit_gaussian <- pls(
  Xtrain,
  Y_multi,
  Xtest,
  Y_multi_test,
  ncomp = 1:2,
  method = "simpls",
  svd.method = "rsvd",
  gaussian_y = TRUE,
  gaussian_y_dim = 3,
  seed = 106
)

dim(fit_gaussian$Ypred)
```

## CUDA Availability

CUDA is optional. Use `has_cuda()` before selecting `backend = "cuda"`.

```{r}
has_cuda()
```

```{r eval = FALSE}
if (has_cuda()) {
  fit_gpu <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "simpls",
    backend = "cuda"
  )

  fit_gpu_lda <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    backend = "cuda",
    classifier = "lda"
  )
}
```

## Helper Functions

`fastcor()` computes fast Pearson-style correlations. `ViP()` returns variable
importance in projection trajectories for fitted PLS models, following the
standard VIP interpretation used for PLS variable ranking (Wold, Sjostrom, and
Eriksson, 2001; Chong and Jun, 2005). These functions are ordinary R-level
helpers and are independent of CUDA availability.

```{r}
C <- fastcor(Xtrain, byrow = FALSE, diag = FALSE)
dim(C)
```

```{r}
vip <- ViP(fit_reg)
dim(vip)
```

## Method References

The implementation in `fastPLS` is not a line-by-line copy of the papers below;
rather, these papers define the statistical algorithms or numerical building
blocks that the package implements and accelerates.

- Baglama, J. and Reichel, L. (2005). Augmented implicitly restarted Lanczos
  bidiagonalization methods. *SIAM Journal on Scientific Computing*, 27,
  19-42.
- Barker, M. and Rayens, W. (2003). Partial least squares for discrimination.
  *Journal of Chemometrics*, 17, 166-173.
- Boulesteix, A.-L. and Strimmer, K. (2007). Partial least squares: a versatile
  tool for the analysis of high-dimensional genomic data. *Briefings in
  Bioinformatics*, 8, 32-44.
- Chong, I.-G. and Jun, C.-H. (2005). Performance of some variable selection
  methods when multicollinearity is present. *Chemometrics and Intelligent
  Laboratory Systems*, 78, 103-112.
- Cover, T. and Hart, P. (1967). Nearest neighbor pattern classification.
  *IEEE Transactions on Information Theory*, 13, 21-27.
- de Jong, S. (1993). SIMPLS: an alternative approach to partial least squares
  regression. *Chemometrics and Intelligent Laboratory Systems*, 18, 251-263.
- Fisher, R. A. (1936). The use of multiple measurements in taxonomic problems.
  *Annals of Eugenics*, 7, 179-188.
- Geladi, P. and Kowalski, B. R. (1986). Partial least-squares regression: a
  tutorial. *Analytica Chimica Acta*, 185, 1-17.
- Halko, N., Martinsson, P.-G., and Tropp, J. A. (2011). Finding structure with
  randomness: probabilistic algorithms for constructing approximate matrix
  decompositions. *SIAM Review*, 53, 217-288.
- Hotelling, H. (1931). The generalization of Student's ratio. *Annals of
  Mathematical Statistics*, 2, 360-378.
- Johnson, W. B. and Lindenstrauss, J. (1984). Extensions of Lipschitz mappings
  into a Hilbert space. *Contemporary Mathematics*, 26, 189-206.
- Mevik, B.-H. and Wehrens, R. (2007). The pls package: principal component and
  partial least squares regression in R. *Journal of Statistical Software*, 18,
  1-23.
- Pearson, K. (1901). On lines and planes of closest fit to systems of points
  in space. *Philosophical Magazine*, 2, 559-572.
- Rosipal, R. and Trejo, L. J. (2001). Kernel partial least squares regression
  in reproducing kernel Hilbert space. *Journal of Machine Learning Research*,
  2, 97-123.
- Shao, Z., Wang, Y., Wang, Q., Jiang, T., Du, Z., Ye, H., Zhuo, D., Chen, Y.,
  and Li, H. (2025). FlashSVD: memory-efficient inference with streaming for
  low-rank models. arXiv:2508.01506.
- Trygg, J. and Wold, S. (2002). Orthogonal projections to latent structures
  (O-PLS). *Journal of Chemometrics*, 16, 119-128.
- Wold, S., Sjostrom, M., and Eriksson, L. (2001). PLS-regression: a basic tool
  of chemometrics. *Chemometrics and Intelligent Laboratory Systems*, 58,
  109-130.
