---
title: "fastPLS Public API and Implementation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastPLS Public API and Implementation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`fastPLS` exposes a compact public API. Users do not need to call the internal
C++, CUDA, or benchmarking helpers directly; all model families and backends are
selected through the user-facing functions below.

| Function | Purpose |
|---|---|
| `pls()` | Fit PLSSVD, SIMPLS, OPLS, or kernel PLS for regression or classification. |
| `predict()` | Predict from fitted `fastPLS`, OPLS, or kernel PLS models. |
| `plot()` | Plot PLS or PCA score maps, with optional confidence or Hotelling T2 ellipses. |
| `pls.single.cv()` | Run fixed-component grouped cross-validation. |
| `optim.pls.cv()` | Choose the best number of components by cross-validation. |
| `pls.double.cv()` | Run nested/double cross-validation with inner component optimization. |
| `fastsvd()` | Run a stand-alone truncated SVD through the bundled backends. |
| `pca()` | Compute PCA using the same SVD backend selector. |
| `fastcor()` | Compute fast Pearson-style correlations. |
| `ViP()` | Compute variable-importance-in-projection trajectories. |
| `has_cuda()` | Check whether CUDA-native fastPLS support is available. |

## Example Data

The examples below use small simulated data so the vignette is fast to build.
Regression responses are supplied as numeric vectors or matrices. Classification
responses are supplied as factors; internally, `fastPLS` handles the response
encoding and returns predicted class labels.

```{r}
library(fastPLS)

set.seed(100)
n <- 80
p <- 12
X <- matrix(rnorm(n * p), nrow = n, ncol = p)

y_signal <- 0.8 * X[, 1] - 0.5 * X[, 2] + rnorm(n, sd = 0.25)
Y_reg <- matrix(y_signal, ncol = 1)
Y_cls <- factor(ifelse(y_signal > median(y_signal), "high", "low"))

test_id <- sample(seq_len(n), 16)
Xtrain <- X[-test_id, , drop = FALSE]
Xtest <- X[test_id, , drop = FALSE]
Ytrain_reg <- Y_reg[-test_id, , drop = FALSE]
Ytest_reg <- Y_reg[test_id, , drop = FALSE]
Ytrain_cls <- Y_cls[-test_id]
Ytest_cls <- Y_cls[test_id]
```

## Fit PLS Models

The `method` argument selects the outer PLS algorithm. The `backend` argument
selects the implementation: `"cpp"` for the compiled CPU implementation and
`"cuda"` for GPU-native execution when available.

The `svd.method` argument selects the CPU singular-vector backend for CPU
models: `"irlba"` or `"cpu_rsvd"`.

```{r}
fit_reg <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:3,
  method = "simpls",
  backend = "cpp",
  svd.method = "irlba",
  fit = TRUE
)

fit_reg$Q2Y
```

The same call can fit the four model families by changing `method`:

- `method = "plssvd"`: direct cross-covariance SVD PLS.
- `method = "simpls"`: accelerated compiled SIMPLS implementation.
- `method = "opls"`: OPLS filtering followed by the selected inner PLS core.
- `method = "kernelpls"`: kernel PLS; the quick-start examples use the linear kernel.

## Implementation of the Main Methods

### Singular-vector backends

The PLS algorithms in `fastPLS` are separated from the inner singular-vector
backend. The compiled CPU implementation supports two truncated SVD routes:
`irlba` and `cpu_rsvd`. The `irlba` backend is a bundled IRLBA-style implicitly
restarted Lanczos bidiagonalization solver. It estimates only the leading
singular triplets needed by the requested PLS components, rather than computing
a full dense SVD. The randomized SVD backend, `cpu_rsvd`, draws a Gaussian test
matrix, forms a low-dimensional sketch of the target matrix, optionally applies
power iterations, orthonormalizes the sketch, and solves the final SVD in the
small projected space. Its accuracy is controlled mainly by `rsvd_oversample`,
`rsvd_power`, and `seed`.

For CUDA builds, GPU execution is selected with `backend = "cuda"`. The removed
hybrid route `svd.method = "cuda_rsvd"` is not part of the public API; instead,
the main matrix operations are performed by GPU-native PLS routines. When the
cross-covariance \(S = X^T Y\) would be large, selected backends can use the
matrix-free `xprod` route. In that route, the code avoids explicitly storing
`S` and evaluates the products required by the SVD through identities such as
\(Sv = X^T(Yv)\) and \(S^Tu = Y^T(Xu)\). For classification problems with many
classes, label-aware products can avoid materializing a dense one-hot response
matrix.

### PLSSVD

`method = "plssvd"` implements the direct cross-covariance SVD formulation of
PLS. After preprocessing, the algorithm computes the dominant singular subspace
of \(X^T Y\) and reuses that subspace to generate the requested component path.
The maximum meaningful number of PLSSVD components is capped by
`min(nrow(X), ncol(X), ncol(Y))`; if a larger value is requested, the package
uses the valid internal rank and records the capped component count.

The main innovation relative to a simple textbook PLSSVD implementation is the
execution strategy. fastPLS can use truncated IRLBA or randomized SVD, compact
low-rank prediction factors, CUDA-native fitting, and the matrix-free `xprod`
route. For large classification tasks, the label-aware PLSSVD path computes
class-wise cross-products from the factor labels and avoids dense one-hot `Y`
unless the legacy dense route is explicitly needed.

### SIMPLS

`method = "simpls"` is the optimized fastPLS SIMPLS core. Statistically, it
keeps the de Jong SIMPLS idea: one supervised latent direction is appended at a
time, the score and loading vectors are computed, an orthogonal loading
direction is constructed, and the cross-covariance is deflated before the next
component.

The implementation differs from a naive classical SIMPLS loop in how it obtains
new singular directions. Instead of always calling the SVD backend for one fresh
vector after every deflation, fastPLS can refresh a small block of candidate
directions from the current deflated cross-covariance. The block is consumed
sequentially: each candidate direction is normalized, converted into a SIMPLS
component, orthogonalized, and followed by the usual SIMPLS deflation before the
next component is accepted. The cached-deflation update explicitly computes
\(v^T S\) and applies the rank-one update \(S <- S - v(v^T S)\), reducing
temporary allocations and repeated projection work. The incremental refresh
uses the previous latent direction as a warm start for the next randomized
refresh, which can reduce the number of iterations needed to recover the next
dominant direction.

These block-refresh, warm-start, and cached-deflation optimizations are
SIMPLS-specific because SIMPLS is a sequential deflation method. PLSSVD uses a
different one-shot cross-covariance SVD strategy. OPLS and kernel PLS inherit
the SIMPLS optimizations only when their `inner.method` is `"simpls"`.

### OPLS

`method = "opls"` first removes response-orthogonal variation from `X` and then
fits the selected inner PLS core to the filtered matrix. The number of
orthogonal components is controlled by `north`. The filtering step estimates
orthogonal scores and loadings, subtracts the corresponding variation from `X`,
and stores the orthogonal filter so the same transformation is applied at
prediction time. After filtering, the model dispatches to either PLSSVD or
SIMPLS through `inner.method`.

### Kernel PLS

`method = "kernelpls"` constructs a kernel representation before applying the
selected inner PLS model. The supported kernels are `"linear"`, `"rbf"`, and
`"poly"`. For nonlinear kernels, the training kernel matrix is centered and the
inner PLS model is fitted in kernel space. For the linear kernel, fastPLS avoids
unnecessary kernel materialization and uses the ordinary linear PLS path.

### Classification heads

For classification, factor responses are treated as PLS-DA targets. The default
classifier, `classifier = "argmax"`, predicts the class with the largest
predicted dummy-response score. This is the classical PLS-DA decision rule.

The LDA heads, `classifier = "lda_cpp"` and `classifier = "lda_cuda"`, fit a
regularized linear discriminant model on the PLS latent scores. For each class,
fastPLS estimates a score-space mean, an empirical prior, and a common pooled
within-class covariance matrix. A small ridge term scaled by the average
covariance diagonal is added for numerical stability. Prediction uses the
standard LDA discriminant score and returns the largest-scoring class. The CUDA
variant performs latent projection and discriminant scoring with native CUDA
kernels when available.

The candidate-kNN heads, `classifier = "candidate_knn_cpp"` and
`classifier = "candidate_knn_cuda"`, are designed for large multiclass PLS-DA.
They first score classes by centroids in supervised PLS score space, keep the
top candidate classes, and then rerank those candidates by same-class nearest
neighbours in the PLS score space. The default tuning parameters are
`candidate_knn_k = 10`, `candidate_tau = 0.2`, `candidate_alpha = 0.75`, and
`candidate_top_m = 20`. `predict()` can also return ranked predictions by using
`top` or `top5 = TRUE`.

```{r}
fit_cls <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 100
)

head(fit_cls$Ypred)
```

For PLS-DA with an LDA prediction head, use the `classifier` argument. The
recommended GPU route is PLSSVD followed by native CUDA LDA:

```{r eval = FALSE}
fit_cls_lda_gpu <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 50,
  method = "plssvd",
  backend = "cuda",
  classifier = "lda_cuda"
)
```

When CUDA is unavailable, the compiled CPU fallback is:

```{r}
fit_cls_lda_cpp <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  classifier = "lda_cpp",
  seed = 100
)

head(fit_cls_lda_cpp$Ypred)
```

## OPLS and Kernel PLS

OPLS and kernel PLS are accessed through the same `pls()` function. The
`inner.method` argument selects whether their internal PLS model uses PLSSVD or
SIMPLS.

```{r}
fit_opls <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:2,
  method = "opls",
  inner.method = "simpls",
  north = 1,
  svd.method = "cpu_rsvd",
  seed = 101
)

class(fit_opls)
```

```{r}
fit_kernel <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "kernelpls",
  kernel = "linear",
  inner.method = "simpls",
  svd.method = "cpu_rsvd",
  seed = 102
)

head(fit_kernel$Ypred)
```

## Prediction

`predict()` applies the preprocessing stored in the fitted object. If `Ytest` is
supplied, predictive metrics are returned when available. The optional
`predict.backend` argument can select the ordinary CPU path, the compact
FlashSVD-style CPU path, or CUDA prediction on CUDA-enabled builds.

```{r}
pred_reg <- predict(fit_reg, Xtest, Ytest = Ytest_reg, proj = TRUE)
names(pred_reg)
dim(pred_reg$Ypred)
```

```{r}
pred_cls <- predict(fit_cls, Xtest, Ytest = Ytest_cls)
table(pred_cls$Ypred[[2]], Ytest_cls)
```

## Score Plots

`plot()` is an S3 method for both fitted PLS models and PCA objects. For PLS
models, refit with `fit = TRUE` to store training scores, or predict with
`proj = TRUE` to store test scores. Ellipses can be ordinary confidence
ellipses or Hotelling T2 ellipses.

```{r fig.width = 5, fig.height = 4}
plot(fit_reg, groups = Ytrain_cls, ellipse = TRUE, ellipse.type = "confidence")
```

The next example uses the two-class `Colon` gene-expression dataset from the
plsgenomics package. Each column is one PLS family, the top row shows training
scores colored by the training class, and the bottom row shows test scores
colored by the predicted class.

```{r colon-binary-pls-plots, fig.width = 11, fig.height = 6}
data("Colon", package = "plsgenomics")
set.seed(200)
X_two <- as.matrix(Colon$X)
y_two <- factor(Colon$Y, labels = c("normal", "tumor"))
train_two <- unlist(tapply(seq_along(y_two), y_two, function(i) {
  sample(i, floor(0.7 * length(i)))
}))

families <- c("plssvd", "simpls", "opls", "kernelpls")
labels <- c("PLSSVD", "SIMPLS", "OPLS", "kernelPLS")
fits_two <- lapply(seq_along(families), function(i) {
  pls(
    X_two[train_two, , drop = FALSE], y_two[train_two],
    X_two[-train_two, , drop = FALSE], y_two[-train_two],
    ncomp = 1:2, method = families[[i]], backend = "cpp",
    svd.method = "cpu_rsvd", fit = TRUE, proj = TRUE,
    north = 1L, kernel = "linear", seed = 200 + i
  )
})

old_par <- par(mfrow = c(2, 4), mar = c(4.2, 4.2, 2.2, 0.8), oma = c(0, 0, 1.2, 0))
for (i in seq_along(families)) {
  plot(fits_two[[i]], score.set = "train", groups = y_two[train_two], ellipse = TRUE, main = paste(labels[[i]], "training"))
}
for (i in seq_along(families)) {
  plot(
    fits_two[[i]],
    score.set = "test",
    groups = fits_two[[i]]$Ypred[[length(fits_two[[i]]$ncomp)]],
    main = paste(labels[[i]], "test prediction")
  )
}
mtext("Two-class colon gene-expression train/test PLS score plots", outer = TRUE, font = 2)
par(old_par)
```

The second example uses the three breast-cancer subtypes in the
`breast.TCGA` multi-omics dataset from mixOmics and fits OPLS with two
orthogonal components (`north = 2`). The example uses the dataset's predefined
training and test sets.

```{r breast-tcga-three-class-opls, fig.width = 7.5, fig.height = 4}
data("breast.TCGA", package = "mixOmics")
X_three <- as.matrix(breast.TCGA$data.train$mrna)
y_three <- droplevels(breast.TCGA$data.train$subtype)
X_three_test <- as.matrix(breast.TCGA$data.test$mrna)
y_three_test <- factor(breast.TCGA$data.test$subtype, levels = levels(y_three))

opls_three <- pls(
  X_three,
  y_three,
  X_three_test,
  y_three_test,
  ncomp = 1:2,
  method = "opls",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  fit = TRUE,
  proj = TRUE,
  north = 2L,
  seed = 201
)

old_par <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 0.8))
plot(
  opls_three,
  score.set = "train",
  groups = y_three,
  ellipse = TRUE,
  ellipse.type = "hotelling",
  main = "OPLS training, north = 2"
)
plot(
  opls_three,
  score.set = "test",
  groups = opls_three$Ypred[[length(opls_three$ncomp)]],
  xlim = c(-0.16, 0.42),
  main = "OPLS test prediction"
)
par(old_par)
```

## Cross-Validation

Use `pls.single.cv()` when the number of components is already fixed. Use
`optim.pls.cv()` to compare a grid of component counts. Use `pls.double.cv()`
for a nested cross-validation workflow. All three helpers use the same compiled
C++/CUDA CV core.

```{r}
cv_fixed <- pls.single.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  ncomp = 2,
  kfold = 3,
  method = "simpls",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 103
)

cv_fixed$metrics
```

```{r}
cv_opt <- optim.pls.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_reg,
  ncomp = 1:3,
  kfold = 3,
  method = "plssvd",
  svd.method = "irlba"
)

cv_opt$optim_comp
```

```{r}
cv_double <- pls.double.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  ncomp = 1:2,
  runn = 1,
  kfold_inner = 3,
  kfold_outer = 3,
  method = "simpls",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 104
)

cv_double$bcomp
```

## PCA and SVD Utilities

`fastsvd()` provides direct access to the same SVD backends used by PLS.
`pca()` builds PCA scores and loadings from those backends.

```{r}
s <- fastsvd(Xtrain, ncomp = 3, method = "cpu_rsvd", seed = 104)
length(s$d)
```

```{r fig.width = 5, fig.height = 4}
pc <- pca(Xtrain, ncomp = 3, svd.method = "cpu_rsvd", seed = 105)
plot(pc, groups = Ytrain_cls, ellipse = TRUE)
```

## Gaussian Response Compression

For high-dimensional responses, `gaussian_y = TRUE` compresses `Y` to a smaller
random Gaussian response space before fitting and decodes predictions back to
the original response scale or class labels. By default, the number of Gaussian
dimensions is `min(ncol(Xtrain), 100)` unless `gaussian_y_dim` is supplied.

```{r}
Y_multi <- cbind(Ytrain_reg, Ytrain_reg^2, sin(Ytrain_reg))
Y_multi_test <- cbind(Ytest_reg, Ytest_reg^2, sin(Ytest_reg))

fit_gaussian <- pls(
  Xtrain,
  Y_multi,
  Xtest,
  Y_multi_test,
  ncomp = 1:2,
  method = "simpls",
  svd.method = "cpu_rsvd",
  gaussian_y = TRUE,
  gaussian_y_dim = 3,
  gaussian_y_seed = 106
)

dim(fit_gaussian$Ypred)
```

## CUDA Availability

CUDA is optional. Use `has_cuda()` before selecting `backend = "cuda"`.

```{r}
has_cuda()
```

```{r eval = FALSE}
if (has_cuda()) {
  fit_gpu <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "simpls",
    backend = "cuda"
  )

  fit_gpu_lda <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    backend = "cuda",
    classifier = "lda_cuda"
  )
}
```

## Helper Functions

`fastcor()` computes fast Pearson-style correlations. `ViP()` returns variable
importance in projection trajectories for fitted PLS models. These functions are
ordinary R-level helpers and are independent of CUDA availability.

```{r}
C <- fastcor(Xtrain, byrow = FALSE, diag = FALSE)
dim(C)
```

```{r}
vip <- ViP(fit_reg)
dim(vip)
```
