---
title: "fastPLS Public API Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastPLS Public API Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`fastPLS` exposes a compact public API. Users do not need to call the internal
C++, CUDA, or benchmarking helpers directly; all model families and backends are
selected through the user-facing functions below.

| Function | Purpose |
|---|---|
| `pls()` | Fit PLSSVD, SIMPLS, OPLS, or kernel PLS for regression or classification. |
| `predict()` | Predict from fitted `fastPLS`, OPLS, or kernel PLS models. |
| `plot()` | Plot PLS or PCA score maps, with optional confidence or Hotelling T2 ellipses. |
| `pls.single.cv()` | Run fixed-component grouped cross-validation. |
| `optim.pls.cv()` | Choose the best number of components by cross-validation. |
| `pls.double.cv()` | Run nested/double cross-validation with inner component optimization. |
| `fastsvd()` | Run a stand-alone truncated SVD through the bundled backends. |
| `pca()` | Compute PCA using the same SVD backend selector. |
| `fastcor()` | Compute fast Pearson-style correlations. |
| `ViP()` | Compute variable-importance-in-projection trajectories. |
| `has_cuda()` | Check whether CUDA-native fastPLS support is available. |

## Example Data

The examples below use small simulated data so the vignette is fast to build.
Regression responses are supplied as numeric vectors or matrices. Classification
responses are supplied as factors; internally, `fastPLS` handles the response
encoding and returns predicted class labels.

```{r}
library(fastPLS)

set.seed(100)
n <- 80
p <- 12
X <- matrix(rnorm(n * p), nrow = n, ncol = p)

y_signal <- 0.8 * X[, 1] - 0.5 * X[, 2] + rnorm(n, sd = 0.25)
Y_reg <- matrix(y_signal, ncol = 1)
Y_cls <- factor(ifelse(y_signal > median(y_signal), "high", "low"))

test_id <- sample(seq_len(n), 16)
Xtrain <- X[-test_id, , drop = FALSE]
Xtest <- X[test_id, , drop = FALSE]
Ytrain_reg <- Y_reg[-test_id, , drop = FALSE]
Ytest_reg <- Y_reg[test_id, , drop = FALSE]
Ytrain_cls <- Y_cls[-test_id]
Ytest_cls <- Y_cls[test_id]
```

## Fit PLS Models

The `method` argument selects the outer PLS algorithm. The `backend` argument
selects the implementation: `"cpp"` for the compiled CPU implementation and
`"cuda"` for GPU-native execution when available.

The `svd.method` argument selects the CPU singular-vector backend for CPU
models: `"irlba"` or `"cpu_rsvd"`.

```{r}
fit_reg <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:3,
  method = "simpls",
  backend = "cpp",
  svd.method = "irlba",
  fit = TRUE
)

fit_reg$Q2Y
```

The same call can fit the four model families by changing `method`:

- `method = "plssvd"`: direct cross-covariance SVD PLS.
- `method = "simpls"`: accelerated compiled SIMPLS implementation.
- `method = "opls"`: OPLS filtering followed by the selected inner PLS core.
- `method = "kernelpls"`: kernel PLS with linear, RBF, or polynomial kernels.

```{r}
fit_cls <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 100
)

head(fit_cls$Ypred)
```

For PLS-DA with an LDA prediction head, use the `classifier` argument. The
recommended GPU route is PLSSVD followed by native CUDA LDA:

```{r eval = FALSE}
fit_cls_lda_gpu <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 50,
  method = "plssvd",
  backend = "cuda",
  classifier = "lda_cuda"
)
```

When CUDA is unavailable, the compiled CPU fallback is:

```{r}
fit_cls_lda_cpp <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  classifier = "lda_cpp",
  seed = 100
)

head(fit_cls_lda_cpp$Ypred)
```

## OPLS and Kernel PLS

OPLS and kernel PLS are accessed through the same `pls()` function. The
`inner.method` argument selects whether their internal PLS model uses PLSSVD or
SIMPLS.

```{r}
fit_opls <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:2,
  method = "opls",
  inner.method = "simpls",
  north = 1,
  svd.method = "cpu_rsvd",
  seed = 101
)

class(fit_opls)
```

```{r}
fit_kernel <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "kernelpls",
  kernel = "rbf",
  inner.method = "simpls",
  svd.method = "cpu_rsvd",
  seed = 102
)

head(fit_kernel$Ypred)
```

## Prediction

`predict()` applies the preprocessing stored in the fitted object. If `Ytest` is
supplied, predictive metrics are returned when available. The optional
`predict.backend` argument can select the ordinary CPU path, the compact
FlashSVD-style CPU path, or CUDA prediction on CUDA-enabled builds.

```{r}
pred_reg <- predict(fit_reg, Xtest, Ytest = Ytest_reg, proj = TRUE)
names(pred_reg)
dim(pred_reg$Ypred)
```

```{r}
pred_cls <- predict(fit_cls, Xtest, Ytest = Ytest_cls)
table(pred_cls$Ypred[[2]], Ytest_cls)
```

## Score Plots

`plot()` is an S3 method for both fitted PLS models and PCA objects. For PLS
models, refit with `fit = TRUE` to store training scores, or predict with
`proj = TRUE` to store test scores. Ellipses can be ordinary confidence
ellipses or Hotelling T2 ellipses.

```{r fig.width = 5, fig.height = 4}
plot(fit_reg, groups = Ytrain_cls, ellipse = TRUE, ellipse.type = "confidence")
```

## Cross-Validation

Use `pls.single.cv()` when the number of components is already fixed. Use
`optim.pls.cv()` to compare a grid of component counts. Use `pls.double.cv()`
for a nested cross-validation workflow. All three helpers use the same compiled
C++/CUDA CV core.

```{r}
cv_fixed <- pls.single.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  ncomp = 2,
  kfold = 3,
  method = "simpls",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 103
)

cv_fixed$metrics
```

```{r}
cv_opt <- optim.pls.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_reg,
  ncomp = 1:3,
  kfold = 3,
  method = "plssvd",
  svd.method = "irlba"
)

cv_opt$optim_comp
```

```{r}
cv_double <- pls.double.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  ncomp = 1:2,
  runn = 1,
  kfold_inner = 3,
  kfold_outer = 3,
  method = "simpls",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 104
)

cv_double$bcomp
```

## PCA and SVD Utilities

`fastsvd()` provides direct access to the same SVD backends used by PLS.
`pca()` builds PCA scores and loadings from those backends.

```{r}
s <- fastsvd(Xtrain, ncomp = 3, method = "cpu_rsvd", seed = 104)
length(s$d)
```

```{r fig.width = 5, fig.height = 4}
pc <- pca(Xtrain, ncomp = 3, svd.method = "cpu_rsvd", seed = 105)
plot(pc, groups = Ytrain_cls, ellipse = TRUE)
```

## Gaussian Response Compression

For high-dimensional responses, `gaussian_y = TRUE` compresses `Y` to a smaller
random Gaussian response space before fitting and decodes predictions back to
the original response scale or class labels. By default, the number of Gaussian
dimensions is `min(ncol(Xtrain), 100)` unless `gaussian_y_dim` is supplied.

```{r}
Y_multi <- cbind(Ytrain_reg, Ytrain_reg^2, sin(Ytrain_reg))
Y_multi_test <- cbind(Ytest_reg, Ytest_reg^2, sin(Ytest_reg))

fit_gaussian <- pls(
  Xtrain,
  Y_multi,
  Xtest,
  Y_multi_test,
  ncomp = 1:2,
  method = "simpls",
  svd.method = "cpu_rsvd",
  gaussian_y = TRUE,
  gaussian_y_dim = 3,
  gaussian_y_seed = 106
)

dim(fit_gaussian$Ypred)
```

## CUDA Availability

CUDA is optional. Use `has_cuda()` before selecting `backend = "cuda"`.

```{r}
has_cuda()
```

```{r eval = FALSE}
if (has_cuda()) {
  fit_gpu <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "simpls",
    backend = "cuda"
  )

  fit_gpu_lda <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    backend = "cuda",
    classifier = "lda_cuda"
  )
}
```

## Helper Functions

`fastcor()` computes fast Pearson-style correlations. `ViP()` returns variable
importance in projection trajectories for fitted PLS models. These functions are
ordinary R-level helpers and are independent of CUDA availability.

```{r}
C <- fastcor(Xtrain, byrow = FALSE, diag = FALSE)
dim(C)
```

```{r}
vip <- ViP(fit_reg)
dim(vip)
```
