---
title: "fastPLS Public API Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastPLS Public API Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`fastPLS` exposes a small public API. Users do not need to call the internal
C++, CUDA, or benchmarking helpers directly. The main entry points are:

- `pls()` for PLSSVD, SIMPLS, OPLS, and kernel PLS.
- `predict()` for predictions from fitted models.
- `pls.single.cv()`, `optim.pls.cv()`, and `pls.double.cv()` for
  cross-validation.
- `fastsvd()` for stand-alone truncated SVD.
- `pca()` and `plot()` for PCA score plots with optional confidence or
  Hotelling's T2 ellipses.
- `fastcor()`, `ViP()`, and `has_cuda()` for common helper tasks.

## Example Data

The examples below use small simulated data so the vignette is fast to build.
Regression responses are supplied as numeric vectors or matrices. Classification
responses are supplied as factors; internally, `fastPLS` handles the response
encoding and returns predicted class labels.

```{r}
library(fastPLS)

set.seed(100)
n <- 80
p <- 12
X <- matrix(rnorm(n * p), nrow = n, ncol = p)

y_signal <- 0.8 * X[, 1] - 0.5 * X[, 2] + rnorm(n, sd = 0.25)
Y_reg <- matrix(y_signal, ncol = 1)
Y_cls <- factor(ifelse(y_signal > median(y_signal), "high", "low"))

test_id <- sample(seq_len(n), 16)
Xtrain <- X[-test_id, , drop = FALSE]
Xtest <- X[test_id, , drop = FALSE]
Ytrain_reg <- Y_reg[-test_id, , drop = FALSE]
Ytest_reg <- Y_reg[test_id, , drop = FALSE]
Ytrain_cls <- Y_cls[-test_id]
Ytest_cls <- Y_cls[test_id]
```

## Fit PLS Models

The `method` argument selects the outer PLS algorithm. The `backend` argument
selects the implementation: `"cpp"` for the compiled CPU implementation,
`"r"` for the pure-R reference implementation, and `"cuda"` for GPU-native
execution when available.

The `svd.method` argument selects the CPU singular-vector backend for CPU
models: `"irlba"`, `"cpu_rsvd"`, or `"exact"`.

```{r}
fit_reg <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:3,
  method = "simpls",
  backend = "cpp",
  svd.method = "irlba",
  fit = TRUE
)

fit_reg$Q2Y
```

```{r}
fit_cls <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 100
)

head(fit_cls$Ypred)
```

For PLS-DA with an LDA prediction head, use the `classifier` argument. The
recommended GPU route is PLSSVD followed by native CUDA LDA:

```{r eval = FALSE}
fit_cls_lda_gpu <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 50,
  method = "plssvd",
  backend = "cuda",
  classifier = "lda_cuda"
)
```

When CUDA is unavailable, the compiled CPU fallback is:

```{r}
fit_cls_lda_cpp <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "plssvd",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  classifier = "lda_cpp",
  seed = 100
)

head(fit_cls_lda_cpp$Ypred)
```

## OPLS and Kernel PLS

OPLS and kernel PLS are accessed through the same `pls()` function. The
`inner.method` argument selects whether their internal PLS model uses PLSSVD or
SIMPLS.

```{r}
fit_opls <- pls(
  Xtrain,
  Ytrain_reg,
  Xtest,
  Ytest_reg,
  ncomp = 1:2,
  method = "opls",
  inner.method = "simpls",
  north = 1,
  svd.method = "cpu_rsvd",
  seed = 101
)

class(fit_opls)
```

```{r}
fit_kernel <- pls(
  Xtrain,
  Ytrain_cls,
  Xtest,
  Ytest_cls,
  ncomp = 1:2,
  method = "kernelpls",
  kernel = "rbf",
  inner.method = "simpls",
  svd.method = "cpu_rsvd",
  seed = 102
)

head(fit_kernel$Ypred)
```

## Prediction

`predict()` applies the preprocessing stored in the fitted object. If `Ytest` is
supplied, predictive metrics are returned when available.

```{r}
pred_reg <- predict(fit_reg, Xtest, Ytest = Ytest_reg, proj = TRUE)
names(pred_reg)
dim(pred_reg$Ypred)
```

```{r}
pred_cls <- predict(fit_cls, Xtest, Ytest = Ytest_cls)
table(pred_cls$Ypred[[2]], Ytest_cls)
```

## Cross-Validation

Use `pls.single.cv()` when the number of components is already fixed. Use
`optim.pls.cv()` to compare a grid of component counts. Use `pls.double.cv()`
for a nested cross-validation workflow.

```{r}
cv_fixed <- pls.single.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_cls,
  ncomp = 2,
  kfold = 3,
  method = "simpls",
  backend = "cpp",
  svd.method = "cpu_rsvd",
  seed = 103
)

cv_fixed$metric
```

```{r}
cv_opt <- optim.pls.cv(
  Xdata = Xtrain,
  Ydata = Ytrain_reg,
  ncomp = 1:3,
  kfold = 3,
  method = "plssvd",
  svd.method = "irlba"
)

cv_opt$optim_comp
```

## PCA and SVD Utilities

`fastsvd()` provides direct access to the same SVD backends used by PLS.
`pca()` builds PCA scores and loadings from those backends.

```{r}
s <- fastsvd(Xtrain, ncomp = 3, method = "cpu_rsvd", seed = 104)
length(s$d)
```

```{r fig.width = 5, fig.height = 4}
pc <- pca(Xtrain, ncomp = 3, svd.method = "cpu_rsvd", seed = 105)
plot(pc, groups = Ytrain_cls, ellipse = TRUE)
```

## Gaussian Response Compression

For high-dimensional responses, `gaussian_y = TRUE` compresses `Y` to a smaller
random Gaussian response space before fitting and decodes predictions back to
the original response scale or class labels. By default, the number of Gaussian
dimensions is `min(ncol(Xtrain), 100)` unless `gaussian_y_dim` is supplied.

```{r}
Y_multi <- cbind(Ytrain_reg, Ytrain_reg^2, sin(Ytrain_reg))
Y_multi_test <- cbind(Ytest_reg, Ytest_reg^2, sin(Ytest_reg))

fit_gaussian <- pls(
  Xtrain,
  Y_multi,
  Xtest,
  Y_multi_test,
  ncomp = 1:2,
  method = "simpls",
  svd.method = "cpu_rsvd",
  gaussian_y = TRUE,
  gaussian_y_dim = 3,
  gaussian_y_seed = 106
)

dim(fit_gaussian$Ypred)
```

## CUDA Availability

CUDA is optional. Use `has_cuda()` before selecting `backend = "cuda"`.

```{r}
has_cuda()
```

```{r eval = FALSE}
if (has_cuda()) {
  fit_gpu <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "simpls",
    backend = "cuda"
  )

  fit_gpu_lda <- pls(
    Xtrain,
    Ytrain_cls,
    Xtest,
    Ytest_cls,
    ncomp = 1:2,
    method = "plssvd",
    backend = "cuda",
    classifier = "lda_cuda"
  )
}
```

## Helper Functions

`fastcor()` computes fast Pearson-style correlations. `ViP()` returns variable
importance in projection trajectories for fitted PLS models.

```{r}
C <- fastcor(Xtrain, byrow = FALSE, diag = FALSE)
dim(C)
```

```{r}
vip <- ViP(fit_reg)
dim(vip)
```
