---
title: "fastPLS: Implementation Details and End-to-End Usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fastPLS: Implementation Details and End-to-End Usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

`fastPLS` implements Partial Least Squares (PLS) for both regression and
classification with:

- two algorithms: `plssvd` and `simpls`
- multiple CPU SVD backends (`irlba`, `cpu_rsvd`) plus separate GPU-native
  wrappers where CUDA is available
- Rcpp and pure-R interfaces (`pls()` and `pls_r()`)
- cross-validation helpers (`optim.pls.cv()`, `pls.double.cv()`)

This vignette explains how each algorithm is implemented and shows complete
usage patterns in both regression and classification contexts.

## Core API

- `pls()`: Rcpp-backed main model fitting function.
- `pls_r()`: pure-R fallback implementation (no CUDA backend).
- `predict.fastPLS()`: prediction for fitted `fastPLS` models.
- `optim.pls.cv()`: component selection with cross-validation.
- `pls.double.cv()`: double cross-validation workflow.
- `svd_methods()`, `svd_run()`, `svd_benchmark()`: SVD backend inspection and benchmarking.
- `transformy()`, `fastcor()`, `ViP()`: helper tools.

## Data Setup

```{r}
library(fastPLS)

set.seed(42)
n <- 90
p <- 12
X <- matrix(rnorm(n * p), nrow = n, ncol = p)
y_reg <- X[, 1] * 0.7 - X[, 2] * 0.4 + rnorm(n, sd = 0.2)
Y_reg <- matrix(y_reg, ncol = 1)
y_cls <- factor(ifelse(y_reg > median(y_reg), "high", "low"))

idx <- sample(seq_len(n), 20)
Xtr <- X[-idx, , drop = FALSE]
Xte <- X[idx, , drop = FALSE]
Ytr_reg <- Y_reg[-idx, , drop = FALSE]
Yte_reg <- Y_reg[idx, , drop = FALSE]
ytr_cls <- y_cls[-idx]
yte_cls <- y_cls[idx]
```

## Implementation Details

### 1) `plssvd`

`plssvd` builds the cross-covariance matrix \(S = X^T Y\), computes a truncated
SVD of `S`, and uses the singular vectors to build latent scores and regression
coefficients.

Important implementation detail:

- the effective rank is capped at  
  `min(nrow(X), ncol(X), ncol(Y))`.
- if a larger `ncomp` is requested, fastPLS warns and uses the capped value.

### 2) `simpls`

`simpls` uses an iterative deflation strategy:

- compute one latent component at a time
- update loadings, scores, and orthogonalized directions
- deflate the cross-covariance structure before the next component

This path supports the same SVD backend selector used by `plssvd`.

### 2b) `simpls_fast` (optional speed-optimized SIMPLS)

`simpls_fast` is an additional optional method exposed in `pls()` that keeps the
SIMPLS-style deflation model but adds performance-oriented controls:

- block latent-direction refresh (`fast_block`)
- optional score centering before normalization (`fast_center_t`)
- optional second orthogonalization pass (`fast_reorth_v`)
- optional incremental block-power warm start (`fast_incremental`, `fast_inc_iters`)
- cached deflation update (`fast_defl_cache`)

These controls are implemented in compiled code and are meant for benchmarking
and high-throughput fitting; default values keep behavior close to standard
SIMPLS while improving throughput on large matrices.

### 3) SVD Backends

- `irlba`: bundled truncated SVD route.
- `cpu_rsvd`: randomized CPU SVD with `rsvd_oversample`, `rsvd_power`, and `seed`.
- GPU execution is exposed through GPU-native wrappers such as `plssvd_gpu()`
  and `simpls_gpu()`, not through `pls(..., svd.method = "cuda_rsvd")`.

### 4) `pls()` vs `pls_r()`

- `pls()` calls compiled code paths for performance.
- `pls_r()` mirrors the same user API in pure R and supports `irlba` and
  `cpu_rsvd`.
- for classification, factor `Y` is converted internally to one-hot encoding and
  class labels are returned by `predict.fastPLS()`.

## Regression Workflow

```{r}
m_reg <- pls(
  Xtr, Ytr_reg,
  ncomp = 1:3,
  method = "simpls",
  svd.method = "irlba",
  fit = TRUE
)

pred_reg <- predict(m_reg, Xte)
str(pred_reg$Ypred)
```

## Fast SIMPLS Workflow

```{r}
m_fast <- pls(
  Xtr, Ytr_reg,
  ncomp = 1:3,
  method = "simpls_fast",
  svd.method = "cpu_rsvd",
  rsvd_oversample = 14L,
  rsvd_power = 1L,
  fast_block = 6L,
  fast_center_t = TRUE,
  fast_reorth_v = FALSE,
  fast_incremental = FALSE,
  fast_inc_iters = 2L,
  fast_defl_cache = TRUE,
  fit = TRUE
)

str(predict(m_fast, Xte)$Ypred)
```

## Classification Workflow

```{r}
m_cls <- pls(
  Xtr, ytr_cls, Xte, yte_cls,
  ncomp = 1:2,
  method = "plssvd",
  svd.method = "irlba",
  fit = TRUE
)

pred_cls <- predict(m_cls, Xte, Ytest = yte_cls)
head(pred_cls$Ypred)
table(pred_cls$Ypred[[2]], yte_cls)
```

## Pure-R Workflow (`pls_r`)

```{r}
m_r <- pls_r(
  Xtr, ytr_cls, Xte, yte_cls,
  ncomp = 1:2,
  method = "simpls",
  svd.method = "cpu_rsvd",
  seed = 123
)

head(m_r$Ypred)
```

## Cross-Validation Workflows

```{r}
cv1 <- optim.pls.cv(
  Xdata = Xtr,
  Ydata = ytr_cls,
  ncomp = 1:3,
  kfold = 3,
  method = "simpls",
  svd.method = "irlba"
)

names(cv1)
```

```{r}
cv2 <- pls.double.cv(
  Xdata = Xtr,
  Ydata = ytr_cls,
  ncomp = 1:3,
  runn = 2,
  kfold_inner = 3,
  kfold_outer = 3,
  method = "simpls",
  svd.method = "irlba"
)

names(cv2)
```

## SVD Utilities

```{r}
svd_methods()
svd_run(Xtr, k = 3, method = "irlba")
svd_benchmark(Xtr, k = 3, methods = c("irlba", "cpu_rsvd"), reps = 2)
```

## Additional Helpers

```{r}
Y_onehot <- transformy(ytr_cls)
dim(Y_onehot)

C <- fastcor(Xtr, byrow = FALSE, diag = FALSE)
dim(C)

vip <- ViP(m_reg)
str(vip)
```

## Practical Notes

- Use `method = "simpls"` when you want robust default behavior across a wide
  range of data shapes.
- Use `method = "simpls_fast"` when you want additional speed controls for
  model-building and are willing to tune numerical tradeoffs.
- Use `plssvd` when you want direct control over rank-constrained latent
  structure; fastPLS automatically enforces the valid component cap.
- For reproducible approximate SVD, keep `svd.method = "cpu_rsvd"` and set
  `seed` (and optionally `rsvd_power`).
- Use `has_cuda()` to check whether GPU-native wrappers are available in your
  session.

## References

- de Jong, S. (1993). SIMPLS: an alternative approach to partial least squares regression. *Chemometrics and Intelligent Laboratory Systems*, 18(3), 251-263.
- Halko, N., Martinsson, P.-G., & Tropp, J. A. (2011). Finding structure with randomness: probabilistic algorithms for constructing approximate matrix decompositions. *SIAM Review*, 53(2), 217-288.
- Musco, C., & Musco, C. (2015). Randomized block Krylov methods for stronger and faster approximate singular value decomposition. arXiv:1504.05477.
